National Longitudinal Study Reveals Growth in Wellbeing Among Those with Spiritual Beliefs

### 2020-11-16 Manuscript.tex
initial commit

---
title             : "National Longitudinal Study Reveals Growth in Wellbeing Among Those with Spiritual Beliefs"
shorttitle        : "Beliefs and Wellbeing"
author: 
  - name          : "Ben Highland"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "Postal address"
    email         : "my@email.com"
    role:         # First author
      - Conceptualization
      - Literature Review
      - Writing - First draft
      - Writing - Review & editing
  - name          : "Ev l Worthing"
    affiliation   : "2"
    address       : "Postal address"
    email         : "my@email.com"
    role:        
      - Writing review & editing
  - name          : "Don E Davis"
    affiliation   : "3"
    address       : "Postal address"
    email         : "my@email.com"
    role:         
      - Writing review & editing
  - name          : "Chris G. Sibley"
    affiliation   : "4"
    address       : "Postal address"
    email         : "my@email.com"
    role:        
      - Writing review & editing
  - name          : "Joseph Bulbulia"
    affiliation   : "5"
    role:         
      - Data analysis
      - Writing second and subsequent drafts
      - Writing review & editing

authornote: |
  Add complete departmental affiliations for each author here. Each new line herein must be indented, like this line.
  Enter author note here.
abstract: |
 Previous research finds an association between spirituality and subjective well-being. However, the widespread use of tautological scales and the reliance on cross-sectional data cast doubts. Here, we leverage nine years of panel response data from a nationally diverse longitudinal study to systematically test whether having spiritual beliefs leads to growth in personal well-being and life satisfaction ($N$ =  20979, New Zealand, 2010-2018). Contrary to previous studies, we find that belief in a spirit or life force predicts initially lower personal wellbeing and life-satisfaction. However, in support of theories that sprituality improves psychological health, beliefs in a spirit or life force predict increasing life satisfaction over time relative to disbelief. These finding holds after adjusting for known demographic influences. Intriguingly these findings also hold among those who believe in a God but disbelieve in a spirit or life force. The causal mechanisms by which beliefs in a spirit or life force increases well-being remain an important horizon for future investigations.

  <!-- https://tinyurl.com/ybremelq -->
keywords          : "Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being"
wordcount         : no
bibliography      : ["b-references.bib"]
floatsintext      : no
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : no
mask              : no
draft             : no
documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
extra_dependencies: ["rotating", "booktabs", "graphicx","longtable"]
header-includes:
   - \usepackage{tcolorbox}
   - \usepackage{paralist}
   - \usepackage{multicol}
---

```{r include = FALSE}
library(knitr)
opts_chunk$set(
    fig.align = "center",
    root.dir = "../"
)
library("papaja")
library("tidyverse")
library("tinytex")
```

```{r load_libraries, echo=FALSE, include=FALSE}
# load libraries
library("styler")
library("tidyverse")
library("lme4")
library("ggplot2")
library("tidyr")
library("sjPlot")
library("ggeffects")
library("Amelia")
library("table1")
library("patchwork")
library("parameters")
library("prettycode")
library("gghighlight")
library("sjPlot")
library("see")
library("viridis")
library("ggpubr")
library("wesanderson")
library("ggsci")
library("here")
library("citr")
#library("furniture")
library("knitr")
library("texreg")
```
```{r load_data, include=FALSE}
ldf.5<-readRDS("~/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/00Bulbulia\ Pubs/2020/ldf.5") 
```
```{r data processing, cache = TRUE, dependson = "ldf.5"}
# filter only those who are measured 
dfben  <-ldf.5 %>%
  dplyr:: filter(Wave !=2009)%>% # exclude wave 1 (no spirit measures that wave)
  dplyr::filter(YearMeasured==1)%>%
  dplyr::group_by(Id) %>% filter(n() > 2)%>% # select those who have responded to at least 3 waves # Also works for 5 and 7
  dplyr::filter(n() !=0)%>%
  dplyr::ungroup(Id)%>%
  dplyr:: mutate(Years = as.numeric(years))%>%
  dplyr::mutate(Age.10yrs = (Age/10))%>%
  dplyr::mutate(Religion.Church.Log = log(Religion.Church+ 1))%>%
  dplyr::mutate(Pol.Orient.S = scale(Pol.Orient, center = TRUE, scale = TRUE),
         Age.10yrs.C = scale(Age.10yrs, center = TRUE, scale = FALSE),
         Male = factor(Gender),
         Employed = factor(Employed),
         Partner = factor(Partner),
         Edu.S = scale(as.numeric(Edu) ,scale=TRUE,center=TRUE),
         Education = as.numeric(Edu),
         Urban = factor(Urban),
         Believe.Spirit = as.factor(Believe.Spirit),
         Believe.God = as.factor(Believe.God),
         EthnicCategories = factor(EthnicCats),
         LIFESAT.S = scale(LIFESAT, center = TRUE, scale = TRUE),
         PWI.S = scale(PWI, center = TRUE, scale = TRUE),
         Deprivation.S = scale(NZdep, scale=TRUE, center=TRUE),
         Relid.C = scale(Relid, scale = TRUE,center=TRUE),
         Religious = as.factor(Religious),
         Religion.Church.Log.C = scale(Religion.Church.Log, center = TRUE, scale = FALSE))
```
```{r}
# nice labels
table1::label(dfben$Age.10yrs.C) <- "Age in Decades (C)"
table1::label(dfben$PWI) <- "Personal Wellbeing"
table1::label(dfben$LIFESAT) <- "Life Satisfaction"

table1::label(dfben$Deprivation.S) <- "Deprivation (S)"
table1::label(dfben$Pol.Orient) <- "Political Conservative"
table1::label(dfben$Pol.Orient.S) <- "Political Conservative (S)"
table1::label(dfben$NZdep) <- "NZ Deprivation Index"
dfben$Partner <- factor(dfben$Partner, labels = c("No Partner","Has Partner"))
dfben$Employed <- factor(dfben$Employed, labels = c("Not Employed","Employed"))
table1::label(dfben$Edu.S) <- "Education (S)"

# create beliefs indicator
dfben$Beliefs<- factor(ifelse(dfben$Believe.God == "Not Believe God" & dfben$Believe.Spirit == "Not Believe Spirit", "_Skeptic_", 
                                        ifelse(dfben$Believe.God == "Not Believe God" & dfben$Believe.Spirit == "Believe Spirit", "_SpiritExcludesGod_",
                                               ifelse(dfben$Believe.God == "Believe God" & dfben$Believe.Spirit == "Believe Spirit","GodAndSpirit","GodExcludesSpirit"))))
```

The relationship between spirituality and psychological well-being is a matter of enduring human fascination [@Ellsworth2010-yu]. Previous attempts to quantify the relationship between these two domains find a positiven association between spirituality and well-being [@Russinova2002-rq; @Ginsburg1995-jr; @Zaza2005-ac; @Koenig2010-gk; @Smith2003-re; @Smith2003-re; @Koenig2001-ow]. However, critics have raised three challenges.

<!-- In spirituality/psychological wellbeing research,sSsubjective in spirituality health research, psychological well-being is typically decomposed into two categories: emotional well-being (i.e., the presence or absence of positive feelings about life) and functional well-being (i.e., psychological and social well-being) [@Keyes2002-bb]. -->

First, previous research has frequently operationalised spirituality using measures of well-being. For example, The Daily Spiritual Experience Scale, which is widely used in spiritual mental health research includes items such as "I feel thankful for my blessings" and "I feel deep inner peace or harmony" to measure spirituality [@Underwood2002-hg]. As Koenig observes, it is unsurprising that well-being predicts well-being –. what Koenig describes as a  "tautology" in spirituality/mental health research [@Koenig2008-lv]. Indeed, recent systematic-review of spirituality/mental health research found that nearly 45% of previous studies investigating spirituality and emotional well-being operationalise spirituality scales using constructs for positive emotional well-being [@Garssen2016-kb]. The first challenge, then, is to define measures of spirituality that are not well-being measures reclocked in other words.

How, then, should researchers operationalise spirituality. We can look to folk traditions for guidance. The term "spirituality" was first used to describe the religious practices of ascetics and monks. more recently, however, the term has taken on a family of broader meaning [@Swinton2001-vr]. Religion is typically operationalized to mean beliefs and practices respecting a God or gods (JB). A recurring element in the use of ‘spirituality,’ as it is distinguished from religion, is belief in a spirit or life force without commitment to a belief in a God or gods [@Hill2000-cl; @Davis2015-tx]. However, the term "spirituality" overlaps with the uses of "religion" in the sense that God or gods are instances of beliefs in spirits – that is one may be both religious and spiritual. Here we operationalize religious belief as either belief in a God or belief in both a God and a spirit or life force. We operationalize spiritual belief as either belief in a spirit or life force without any corresponding belief in a God. However, the term "spirituality" overlaps with the uses of "religion" in the sense that God or gods are instances of beliefs in spirits – that is one may be both religious and spiritual. Here we operationalize religious belief as either belief in a God or belief in both a God and a spirit or life force.  isNotably, previous spirituality mental health research has been disproportionately conducted on North American samples. In the United States, especially, samples include a majority of people who believe in both a God as well as and life forces other than God [@Smith2003-re; @Koenig2010-gk]. Although such studies drawing from such samples mightmay clarify the effects of religious spirituality on psychological subjective well-being, the question of how non-religious spirituality affects psychological well-being that is – or spiritual beliefs without any belief in a God or gods – affects subjective well-being remain unclear.  (For discussion, see: Therefore, to better understand possible distinctions between religious spirituality and non-religious spirituality for subjective well-being outcomes, samples collected from cultures outside the North Atlantic that contain a more equal ratio of the two are needed [@Garssen2016-km]).

Third, most research investigating the relationship between spirituality and subjective well-being has relied on cross-sectional samples. These studies include high-quality reviews and meta-analyses [@Ano2005-hx; @Hackney2003-rs; @Sawatzky2005-rw; @Smith2003-re; @Visser2010-kq; @Yonker2012-zg]. However, to assess whetherto whether spiritual beliefs lead to a growth, stability, or decline in psychological well-being relative to people who lack such beliefs , including people who believe in God but not a spirit or lifeforce, requires repeatedly measuring the same individuals over time. clarify causal effects requires measuring the same individuals over time and assessing how life turns out for demographically similar people who differ in their spiritual and religious beliefs over their lifespans  and orientations [@Garssen2016-km].

Our study addresses these three limitations in previous research.improves on past methods the following ways: First, we avoid a tautological elision of "spirituality" and "well-being"y by adopting a non-affective measure of spirituality through use of general, non-affective measures: "Do you believe in some form of a spirit or life force?" Our measures of psychological well being are [BEN SAY HERE] and Second we adopt a four-level categorical indicator that disentangles spiritual and religious beliefs: (1) Skeptics: those who neither believe in a God nor a life force; (2) SpiritExcludeGod: those who no not believe in a God  but believe in a spirit or life force; (3) GodAndSpirit: those who believe in a God and spirit; (4) GodExcludesSpirit: those who believe in both a God and a spirit or life force "Do you believe in a god?". Second, we address the prior dependence on cross-sectional samples taken from the United States by analyzing data taken from By sampling from respondents in New Zealand, a country with [GIVE NON_RELIGIOUS SPIRITUAL STATS HERE).  Third, we are able to assess whether spirit beliefs are associate with a growth in psychological wellbeing address causation by weusing we use longitudinal responses from athis nationally diverse probability sample year over an nine-yearseight-year period with nine measurement points  to assess whether spirit beliefs lead to growth, stability, or decline in psychological well-beingwellbeing (2010 to 2018 New Zealand Attitudes and Values Study, NZAVS).

# Method

<!-- We rigorously assess the relationship between our non-affective measures and two constructs of subjective well-being (personal well-being and life satisfaction), adjusting for known demographic and personal influences, including age, sex, deprivation, and political orientation (Okulicz-Kozaryn, Holmes, and Avery 2014; Cuesta and Budría 2012; Røysamb et al. 2002; Hansen and Slagsvold 2012; Chen and van Ours, n.d.; Vaughan et al. 1985; Easterlin, Angelescu, and Zweig 2011; Stam et al. 2016; Yakovlev and Leguizamon 2012). Collectively, then, our approach avoids the limitations of many previous studies and helps to clarify existing literature concerning how best to study the relationship between spirituality and subjective well-being. -->

## Ethics Statement

@@

## Sampling Procedure

The NZAVS is an annual, longitudinal national probability sample of registered New Zealand voters, which was started in 2009. The Time 10 wave of the NZAVS contained responses from 47,951 participants (17,981 retained from one or more previous waves and 29,970 new additions from booster sampling and/or unmatched participants or unsolicited opt-ins). Participants who provided an email address were also emailed and invited to complete an online version if they preferred. We offered a prize draw for participation, non-respondents were emailed and phoned multiple times, and all participants were mailed a Season’s Greetings card from the NZAVS research team and informed that they had been automatically entered into a bonus seasonal grocery voucher prize draw. We also mailed our yearly pamphlet summarizing key research findings published during the current wave of the study.

## Participants

@@

## Measures

**Life Satisfaction** Life satisfaction is a measure of emotional well-being that was assessed using a 2-item version of the Satisfaction With Life Scale, which has previously been shown to correlate with aspects of religiosity [@Diener1985-xy]. Participants rated their agreement with the statements (a) "I am satisfied with my life"; and (b) "In most ways my life is close to ideal." The items were rated on 7-point response options ranging from 1 = strongly disagree to 7 = strongly agree. The mean Cronbach’s alpha for this 2-item scale was $\alpha$ = 0.90. Higher scores on this scale indicate higher life satisfaction. Overall scores were means of the 2-items. For detailed information pertaining to means, standard deviations, and missingness, see Table 1.

**Personal well-being** Personal well-being is a measure of function all well-being that was assessed using a 4-item version of the Australian Unity well-being Index (Cummins et al. 2017). Participants rated their satisfaction with (a) "Your standard of living"; (b) "Your health"; (c) "Your future security"; (d) "Your personal relationships." Items were rated using a 10-point response option ranging from 1 = completely dissatisfied to 10 = completely satisfied. The mean Cronbach’s alpha for this scale was $\alpha$ = 0.90. Higher scores on this scale indicate higher personal well-being. Overall scores were means of the 4-items. For detailed information pertaining to means, standard deviations, and missingness, see Table 1.

***Belief in a Spirit or a God***  We assess belief in a spirit or a God by asking two non-affective questions: "Do you believe in a spirit or life force" and "Do you believe in a god?" Responses were coded as "Belief in a spirit that excludes a god" for those who indicated a belief in both a spirit or life force and a god, "Belief that includes a god" for those who indicated a belief in a god with or without an accompanying belief in a spirit or life force, and "skeptics" for those who indicated belief in neither a spirit or life force nor a god. These items appear in the NZAVS questionnaire and were developed for the 2005 Eurobarometer. For detailed information pertaining to beliefs, see Table 1.

@@

## Demographic Indicators

**Age** Age was put into units of 10 years and centered at its mean. For detailed information pertaining to yearly means, standard deviations, and missingness, see Table 2.
Gender. Gender was assessed by asking participants if they were "Male" was coded as "1" and "Female" was coded as "0." For detailed information pertaining to yearly responses and missingness, see Table 1.

**Education**  Education level was measured using an 11-point rating developed by the New Zealand Qualification Authority known as the New Zealand Qualification Framework (NZQF; 0 = no qualification, 10 = doctoral degree). Education was centered at its mean and standardized. For detailed information pertaining to yearly means, standard deviations, and missingness, see Table 1.

**Deprivation** We measured the socio-economic status of participants’ immediate (small area) neighborhood using the 2013 New Zealand Deprivation Index, which uses aggregate census information about the residents of small neighborhood-type units to assign a decile-rank index from 1 (most affluent) to 10 (most impoverished) [@Atkinson2014-ex]. The index is based on a Principal Components Analysis of the following nine variables (in weighted order): the proportion of adults who received a means-tested benefit, household income, proportion not owning own home, proportion single-parent families, proportion unemployed, proportion lacking qualifications, proportion household crowding, proportion no telephone access, and proportion no car access. Thus, the index reflects the average level of deprivation for small neighborhood-type units (or small community areas of about 80–90 people each) across the entire country. Our sample had a mean deprivation index of 4.80 (SD = 2.79). Deprivation was centered at its mean and standardized. For detailed information pertaining to yearly means, standard deviations, and missingness, see Table 1.

**Employed** Employment status was assessed by asking participants if they were currently working, "yes" was coded as "1" and "no" was coded as "0." For detailed information pertaining to yearly responses and missingness, see Table 1.

**Partner** Participants were asked if they were in a relationship, "yes" was coded as "1" and no was coded as "0." For detailed information pertaining to yearly responses and missingness, see Table 1.

**Ethnicity** Ethnicity was assessed using four basic categories: (1) New Zealand European/Pakeha, (2) Maori, (3) Pacific Islander, and (4) Asian. For detailed information pertaining to yearly responses and missingness, see Table 1.

**Urban** People were coded as either residing in an urban "1" or rural "0" area based on New Zealand census data. For detailed information pertaining to yearly responses and missingness, see Table 1.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  english,
  man]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={National Longitudinal Study Reveals Growth in Wellbeing Among Those with Spiritual Beliefs},
  pdfauthor={Ben Highland1, Ev l Worthing2, Don E Davis3, Chris G. Sibley4, \& Joseph Bulbulia5},
  pdflang={en-EN},
  pdfkeywords={Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

% Table formatting
\usepackage{longtable}
\usepackage{lscape}
% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
\usepackage{multirow}		% Table styling
\usepackage{tabularx}		% Control Column width
\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
\usepackage{threeparttablex}            % Lets threeparttable work with longtable

% Create new environments so endfloat can handle them
% \newenvironment{ltable}
%   {\begin{landscape}\begin{center}\begin{threeparttable}}
%   {\end{threeparttable}\end{center}\end{landscape}}
\newenvironment{lltable}{\begin{landscape}\begin{center}\begin{ThreePartTable}}{\end{ThreePartTable}\end{center}\end{landscape}}

% Enables adjusting longtable caption width to table width
% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
\makeatletter
\newcommand\LastLTentrywidth{1em}
\newlength\longtablewidth
\setlength{\longtablewidth}{1in}
\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}

% \setlength{\parindent}{0.5in}
% \setlength{\parskip}{0pt plus 0pt minus 0pt}

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\HyOrg@maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother
\shorttitle{Beliefs and Wellbeing}
\keywords{Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being\newline\indent Word count: FALSE}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{csquotes}
\usepackage{tcolorbox}
\usepackage{paralist}
\usepackage{multicol}
\ifxetex
  % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
  \usepackage{polyglossia}
  \setmainlanguage[]{english}
\else
  \usepackage[shorthands=off,main=english]{babel}
\fi

\title{National Longitudinal Study Reveals Growth in Wellbeing Among Those with Spiritual Beliefs}
\author{Ben Highland\textsuperscript{1}, Ev l Worthing\textsuperscript{2}, Don E Davis\textsuperscript{3}, Chris G. Sibley\textsuperscript{4}, \& Joseph Bulbulia\textsuperscript{5}}
\date{}

\authornote{

Add complete departmental affiliations for each author here. Each new line herein must be indented, like this line.
Enter author note here.

The authors made the following contributions. Ben Highland: Conceptualization, Literature Review, Writing - First draft, Writing - Review \& editing; Ev l Worthing: Writing review \& editing; Don E Davis: Writing review \& editing; Chris G. Sibley: Writing review \& editing; Joseph Bulbulia: Data analysis, Writing second and subsequent drafts, Writing review \& editing.

Correspondence concerning this article should be addressed to Ben Highland, Postal address. E-mail: \href{mailto:my@email.com}{\nolinkurl{my@email.com}}

\affiliation{\phantom{0}}

\abstract{
Previous research finds an association between spirituality and subjective well-being. However, the widespread use of tautological scales and the reliance on cross-sectional data cast doubts. Here, we leverage nine years of panel response data from a nationally diverse longitudinal study to systematically test whether having spiritual beliefs leads to growth in personal well-being and life satisfaction (\(N\) = 20979, New Zealand, 2010-2018). Contrary to previous studies, we find that belief in a spirit or life force predicts initially lower personal wellbeing and life-satisfaction. However, in support of theories that sprituality improves psychological health, beliefs in a spirit or life force predict increasing life satisfaction over time relative to disbelief. These finding holds after adjusting for known demographic influences. Intriguingly these findings also hold among those who believe in a God but disbelieve in a spirit or life force. The causal mechanisms by which beliefs in a spirit or life force increases well-being remain an important horizon for future investigations.
}

\begin{document}
\maketitle

The relationship between spirituality and psychological well-being is a matter of enduring human fascination (Ellsworth \& Ellsworth, 2010). Previous attempts to quantify the relationship between these two domains find a positiven association between spirituality and well-being (Ginsburg, Quirt, Ginsburg, \& MacKillop, 1995; Koenig, 2010; Koenig, Associate Professor of Psychology and Religious Studies Michael E McCullough, PhD, McCullough, Larson, \& Adjunct Professor of Psychiatry David B Larson, 2001; Russinova, Wewiorski, \& Cash, 2002; Smith, McCullough, \& Poll, 2003; Smith et al., 2003; Zaza, Sellick, \& Hillier, 2005). However, critics have raised three challenges.

First, previous research has frequently operationalised spirituality using measures of well-being. For example, The Daily Spiritual Experience Scale, which is widely used in spiritual mental health research includes items such as \enquote{I feel thankful for my blessings} and \enquote{I feel deep inner peace or harmony} to measure spirituality (Underwood \& Teresi, 2002). As Koenig observes, it is unsurprising that well-being predicts well-being --. what Koenig describes as a \enquote{tautology} in spirituality/mental health research (Koenig, 2008). Indeed, recent systematic-review of spirituality/mental health research found that nearly 45\% of previous studies investigating spirituality and emotional well-being operationalise spirituality scales using constructs for positive emotional well-being (Garssen et al., 2016). The first challenge, then, is to define measures of spirituality that are not well-being measures reclocked in other words.

How, then, should researchers operationalise spirituality. We can look to folk traditions for guidance. The term \enquote{spirituality} was first used to describe the religious practices of ascetics and monks. more recently, however, the term has taken on a family of broader meaning (Swinton, 2001). Religion is typically operationalized to mean beliefs and practices respecting a God or gods (JB). A recurring element in the use of \enquote{spirituality,} as it is distinguished from religion, is belief in a spirit or life force without commitment to a belief in a God or gods (Davis et al., 2015; Hill et al., 2000). However, the term \enquote{spirituality} overlaps with the uses of \enquote{religion} in the sense that God or gods are instances of beliefs in spirits -- that is one may be both religious and spiritual. Here we operationalize religious belief as either belief in a God or belief in both a God and a spirit or life force. We operationalize spiritual belief as either belief in a spirit or life force without any corresponding belief in a God. However, the term \enquote{spirituality} overlaps with the uses of \enquote{religion} in the sense that God or gods are instances of beliefs in spirits -- that is one may be both religious and spiritual. Here we operationalize religious belief as either belief in a God or belief in both a God and a spirit or life force. isNotably, previous spirituality mental health research has been disproportionately conducted on North American samples. In the United States, especially, samples include a majority of people who believe in both a God as well as and life forces other than God (Koenig, 2010; Smith et al., 2003). Although such studies drawing from such samples mightmay clarify the effects of religious spirituality on psychological subjective well-being, the question of how non-religious spirituality affects psychological well-being that is -- or spiritual beliefs without any belief in a God or gods -- affects subjective well-being remain unclear. (For discussion, see: Therefore, to better understand possible distinctions between religious spirituality and non-religious spirituality for subjective well-being outcomes, samples collected from cultures outside the North Atlantic that contain a more equal ratio of the two are needed (Garssen \& Visser, 2016)).

Third, most research investigating the relationship between spirituality and subjective well-being has relied on cross-sectional samples. These studies include high-quality reviews and meta-analyses (Ano \& Vasconcelles, 2005; Hackney \& Sanders, 2003; Sawatzky, Ratner, \& Chiu, 2005; Smith et al., 2003; Visser, Garssen, \& Vingerhoets, 2010; Yonker, Schnabelrauch, \& Dehaan, 2012). However, to assess whetherto whether spiritual beliefs lead to a growth, stability, or decline in psychological well-being relative to people who lack such beliefs , including people who believe in God but not a spirit or lifeforce, requires repeatedly measuring the same individuals over time. clarify causal effects requires measuring the same individuals over time and assessing how life turns out for demographically similar people who differ in their spiritual and religious beliefs over their lifespans and orientations (Garssen \& Visser, 2016).

Our study addresses these three limitations in previous research.improves on past methods the following ways: First, we avoid a tautological elision of \enquote{spirituality} and \enquote{well-being}y by adopting a non-affective measure of spirituality through use of general, non-affective measures: \enquote{Do you believe in some form of a spirit or life force?} Our measures of psychological well being are {[}BEN SAY HERE{]} and Second we adopt a four-level categorical indicator that disentangles spiritual and religious beliefs: (1) Skeptics: those who neither believe in a God nor a life force; (2) SpiritExcludeGod: those who no not believe in a God but believe in a spirit or life force; (3) GodAndSpirit: those who believe in a God and spirit; (4) GodExcludesSpirit: those who believe in both a God and a spirit or life force \enquote{Do you believe in a god?}. Second, we address the prior dependence on cross-sectional samples taken from the United States by analyzing data taken from By sampling from respondents in New Zealand, a country with {[}GIVE NON\_RELIGIOUS SPIRITUAL STATS HERE). Third, we are able to assess whether spirit beliefs are associate with a growth in psychological wellbeing address causation by weusing we use longitudinal responses from athis nationally diverse probability sample year over an nine-yearseight-year period with nine measurement points to assess whether spirit beliefs lead to growth, stability, or decline in psychological well-beingwellbeing (2010 to 2018 New Zealand Attitudes and Values Study, NZAVS).

\hypertarget{method}{%
\section{Method}\label{method}}

\hypertarget{ethics-statement}{%
\subsection{Ethics Statement}\label{ethics-statement}}

@@

\hypertarget{sampling-procedure}{%
\subsection{Sampling Procedure}\label{sampling-procedure}}

The NZAVS is an annual, longitudinal national probability sample of registered New Zealand voters, which was started in 2009. The Time 10 wave of the NZAVS contained responses from 47,951 participants (17,981 retained from one or more previous waves and 29,970 new additions from booster sampling and/or unmatched participants or unsolicited opt-ins). Participants who provided an email address were also emailed and invited to complete an online version if they preferred. We offered a prize draw for participation, non-respondents were emailed and phoned multiple times, and all participants were mailed a Season's Greetings card from the NZAVS research team and informed that they had been automatically entered into a bonus seasonal grocery voucher prize draw. We also mailed our yearly pamphlet summarizing key research findings published during the current wave of the study.

\hypertarget{participants}{%
\subsection{Participants}\label{participants}}

@@

\hypertarget{measures}{%
\subsection{Measures}\label{measures}}

\textbf{Life Satisfaction} Life satisfaction is a measure of emotional well-being that was assessed using a 2-item version of the Satisfaction With Life Scale, which has previously been shown to correlate with aspects of religiosity (Diener, Emmons, Larsen, \& Griffin, 1985). Participants rated their agreement with the statements (a) \enquote{I am satisfied with my life}; and (b) \enquote{In most ways my life is close to ideal.} The items were rated on 7-point response options ranging from 1 = strongly disagree to 7 = strongly agree. The mean Cronbach's alpha for this 2-item scale was \(\alpha\) = 0.90. Higher scores on this scale indicate higher life satisfaction. Overall scores were means of the 2-items. For detailed information pertaining to means, standard deviations, and missingness, see Table 1.

\textbf{Personal well-being} Personal well-being is a measure of function all well-being that was assessed using a 4-item version of the Australian Unity well-being Index (Cummins et al.~2017). Participants rated their satisfaction with (a) \enquote{Your standard of living}; (b) \enquote{Your health}; (c) \enquote{Your future security}; (d) \enquote{Your personal relationships.} Items were rated using a 10-point response option ranging from 1 = completely dissatisfied to 10 = completely satisfied. The mean Cronbach's alpha for this scale was \(\alpha\) = 0.90. Higher scores on this scale indicate higher personal well-being. Overall scores were means of the 4-items. For detailed information pertaining to means, standard deviations, and missingness, see Table 1.

\textbf{\emph{Belief in a Spirit or a God}} We assess belief in a spirit or a God by asking two non-affective questions: \enquote{Do you believe in a spirit or life force} and \enquote{Do you believe in a god?} Responses were coded as \enquote{Belief in a spirit that excludes a god} for those who indicated a belief in both a spirit or life force and a god, \enquote{Belief that includes a god} for those who indicated a belief in a god with or without an accompanying belief in a spirit or life force, and \enquote{skeptics} for those who indicated belief in neither a spirit or life force nor a god. These items appear in the NZAVS questionnaire and were developed for the 2005 Eurobarometer. For detailed information pertaining to beliefs, see Table 1.

@@

\hypertarget{demographic-indicators}{%
\subsection{Demographic Indicators}\label{demographic-indicators}}

\textbf{Age} Age was put into units of 10 years and centered at its mean. For detailed information pertaining to yearly means, standard deviations, and missingness, see Table 2.
Gender. Gender was assessed by asking participants if they were \enquote{Male} was coded as \enquote{1} and \enquote{Female} was coded as \enquote{0.} For detailed information pertaining to yearly responses and missingness, see Table 1.

\textbf{Education} Education level was measured using an 11-point rating developed by the New Zealand Qualification Authority known as the New Zealand Qualification Framework (NZQF; 0 = no qualification, 10 = doctoral degree). Education was centered at its mean and standardized. For detailed information pertaining to yearly means, standard deviations, and missingness, see Table 1.

\textbf{Deprivation} We measured the socio-economic status of participants' immediate (small area) neighborhood using the 2013 New Zealand Deprivation Index, which uses aggregate census information about the residents of small neighborhood-type units to assign a decile-rank index from 1 (most affluent) to 10 (most impoverished) (Atkinson, Salmond, \& Crampton, 2014). The index is based on a Principal Components Analysis of the following nine variables (in weighted order): the proportion of adults who received a means-tested benefit, household income, proportion not owning own home, proportion single-parent families, proportion unemployed, proportion lacking qualifications, proportion household crowding, proportion no telephone access, and proportion no car access. Thus, the index reflects the average level of deprivation for small neighborhood-type units (or small community areas of about 80--90 people each) across the entire country. Our sample had a mean deprivation index of 4.80 (SD = 2.79). Deprivation was centered at its mean and standardized. For detailed information pertaining to yearly means, standard deviations, and missingness, see Table 1.

\textbf{Employed} Employment status was assessed by asking participants if they were currently working, \enquote{yes} was coded as \enquote{1} and \enquote{no} was coded as \enquote{0.} For detailed information pertaining to yearly responses and missingness, see Table 1.

\textbf{Partner} Participants were asked if they were in a relationship, \enquote{yes} was coded as \enquote{1} and no was coded as \enquote{0.} For detailed information pertaining to yearly responses and missingness, see Table 1.

\textbf{Ethnicity} Ethnicity was assessed using four basic categories: (1) New Zealand European/Pakeha, (2) Maori, (3) Pacific Islander, and (4) Asian. For detailed information pertaining to yearly responses and missingness, see Table 1.
